%==============================================================================
% Sjabloon onderzoeksvoorstel bachproef
%==============================================================================
% Gebaseerd op document class `hogent-article'
% zie <https://github.com/HoGentTIN/latex-hogent-article>

% Voor een voorstel in het Engels: voeg de documentclass-optie [english] toe.
% Let op: kan enkel na toestemming van de bachelorproefcoördinator!
\documentclass{hogent-article}[english]
\usepackage[backend=biber]{biblatex}

% Invoegen bibliografiebestand
\addbibresource{voorstel.bib}

% Informatie over de opleiding, het vak en soort opdracht
% \studyprogramme{Professionele bachelor toegepaste informatica}
% \course{Bachelorproef}
% \assignmenttype{Onderzoeksvoorstel}
% Voor een voorstel in het Engels, haal de volgende 3 regels uit commentaar
\studyprogramme{Bachelor of applied information technology}
\course{Bachelor thesis}
\assignmenttype{Research proposal}

\academicyear{2024-2025} % TODO: pas het academiejaar aan

% TODO: Werktitel
\title{Leveraging Machine Learning to Examine the Link Between Bitcoin and Tezos Market Dynamics.}

% TODO: Studentnaam en emailadres invullen
\author{Hanno van Baarle}
\email{hanno.vanbaarle@student.hogent.be}

% TODO: Medestudent
% Gaat het om een bachelorproef in samenwerking met een student in een andere
% opleiding? Geef dan de naam en emailadres hier
% \author{Yasmine Alaoui (naam opleiding)}
% \email{yasmine.alaoui@student.hogent.be}

% TODO: Geef de co-promotor op
\supervisor[Co-promotor]{F. J. Kirk}

% Binnen welke specialisatierichting uit 3TI situeert dit onderzoek zich?
% Kies uit deze lijst:
%
% - Mobile \& Enterprise development
% - AI \& Data Engineering
% - Functional \& Business Analysis
% - System \& Network Administrator
% - Mainframe Expert
% - Als het onderzoek niet past binnen een van deze domeinen specifieer je deze
%   zelf
%
\specialisation{AI \& Data Engineering}
\keywords{Bitcoin, Tezos, Machine Learning, Price Prediction, Cryptocurrency, Blockchain}

\begin{document}

\begin{abstract}
  This thesis investigates the correlation between Bitcoin and Tezos price movements and explores the application of machine learning techniques to predict Tezos trends. 
  The motivation for this study stems from operational challenges faced by decentralized exchanges like CrunchySwap, which rely on Tezos (\$XTZ) for gas fees. Sudden price fluctuations can lead to inefficiencies and higher user costs. 
  Given Bitcoin's outsized influence on the broader cryptocurrency market, this research examines whether its price dynamics can serve as a predictive signal for Tezos.
  To address this, historical OHLCV data for Bitcoin and Tezos over the past 3–4 years were collected from Binance. 
  The data were cleaned, engineered into additional features (e.g., moving averages, volatility), and standardized before being fed into three types of models: linear regression, random forest, and long short-term memory (LSTM) networks. These models were evaluated using metrics such as MAE, MSE, and R² to assess their predictive accuracy.
  The expected findings suggest a strong, potentially time-lagged correlation between Bitcoin and Tezos, with Tezos often reacting to Bitcoin’s movements with some delay. Such a pattern, if confirmed, could enable exchanges to anticipate gas price changes and adjust fee structures proactively. 
  The results would imply that Bitcoin serves not just as a market leader but also as a practical forecasting tool for altcoin ecosystems.
\end{abstract}

\tableofcontents

% De hoofdtekst van het voorstel zit in een apart bestand, zodat het makkelijk
% kan opgenomen worden in de bijlagen van de bachelorproef zelf.
\input{voorstel-inhoud}

\printbibliography[heading=bibintoc]

\end{document}